IA Random Choice SPH Scheme with Adaptive Viscosity

### 2018-06-01 ebp-JCPManuscript.tex
ap_clean edits1

\modulolinenumbers[5]
\journal{Journal of Computational Physics}
%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%
%% Numbered
\bibliographystyle{model1-num-names}
%% Numbered without titles
%\bibliographystyle{model1a-num-names}
%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}
%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}
%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}
%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{frontmatter}
\title{Integrating random choice method with SPH : a new scheme with adaptive artificial viscosity and shock capturing ability}
%\tnotetext[mytitlenote]{Can add  title note here}
%% Group authors per affiliation:
\author{Zhixuan Cao}
\author{Abani Patra \fnref{mycorrespondingauthor}}
\address{Department of Mechanical and Aerospace Engineering,The State University of New York at Buffalo, Amherst, NY, 14260-4200, USA}
\author{E. Bruce Pitman}
\address{Department of Materials Design and Innovation, The State University of New York at Buffalo, Amherst, NY, 14260-4200, USA}
%\fntext[myfootnote]{Since 1880.}
%% or include affiliations in footnotes:
%\author[mymainaddress,mysecondaryaddress]%{Elsevier Inc}
%\ead[url]{www.elsevier.com}
%\author[mysecondaryaddress]{Global Customer Service\corref{mycorrespondingauthor}}
\fntext[mycorrespondingauthor]{abani@buffalo.edu}
%\cortext[mycorrespondingauthor]{abani@buffalo.edu}
%\ead{abani@buffalo.edu}
%\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
%\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
Standard smoothed particle hydrodynamics (SPH) employs an artificial viscosity, defined explicitly by constant coefficients, to properly capture hydrodynamic shock waves.
Such methodology usually provides dissipation more than need and smears the discontinuities if the artificial viscosity coefficients are not tuned.
Excessively large artificial viscosity might also damp generation of turbulence by fluid instability.
We present in this article a novel SPH scheme with adaptive artificial viscosity, in which the artificial viscosity is determined implicitly based on solution of a local Riemann problem. Different from Godunov SPH (GSPH) method, which utilizing solution of local Riemann problem at an imaginary interface, the new method randomly sample the solution of Riemann problem mimicking the random choice method (RCM). One dimensional shock tube tests demonstrate several attractive features of this new scheme. First of all, it introduces less but sufficient artificial dissipation for shock capturing compared with classical SPH and GSPH. The pressure ``wiggle" around contact discontinuity, which persists in SPH results, is ameliorated by the new scheme. The new method also shows higher order of convergence than GSPH.
%when the same construction of local Riemann problem and Riemann solvers are adopted by both methods. 
Classical RCM method would lose its desirable property of low numerical dissipation when solving multi-dimensional unsteady non-linear systems. Combining of RCM with SPH overcomes the shortcoming  of classical RCM method. Three dimensional time dependent compressible Euler equations are solved by the new scheme for simulating a free jet flow, demonstrating that the desirable less dissipation property of the new scheme persists in multi-dimensional applications.
\end{abstract}
\begin{keyword}
Smoothed Particle Hydrodynamics (SPH), Random Choice Method (RCM), Riemann Solver, Artificial viscosity, Hyperbolic PDEs
\end{keyword}
\end{frontmatter}

\linenumbers
\section{Introduction}
Smoothed particle hydrodynamics (SPH) \citep{gingold1977smoothed,lucy1977numerical} is a mesh-free particle method based on Lagrangian formulation and has been widely applied to different areas in engineering and science. SPH has both advantages and disadvantages with respect to other numerical techniques. One main advantage is that it automatically adapts to following dynamic flows and arbitrary geometries. Another advantage is that it needs less programming effort to include more physics. In addition, SPH particles only interact with neighboring particles, which avoids massive global communication in parallel implementation of SPH.
Historically, SPH has difficulty in capturing discontinuities, such as shocks or contact discontinuities. Inviscid description of fluids becomes invalid at shock fronts, where specific entropy of the fluids increases through the conversion of mechanical energy into internal energy. Most SPH codes have included an artificial viscosity term explicitly, both for stabilizing the simulation and for handling shock discontinuities by dissipating local velocity differences and converting them into heat \citep{monaghan1983shock, monaghan1997sph, klapp2012strong}. Without knowing the minimum amount of dissipation for shock capturing, such methodology usually provides dissipation more than need and smears the discontinuity if the artificial viscosity coefficients are not tuned. The method of using constant artificial viscosity might also introduce spurious effects at region away from shocks, for instance, causing an unphysical damping of turbulent mixing (as shown, for example, by \citet{borgani2012hydrodynamic}) or spurious
shearing torques in rotating flows \citep{flebbe1994smoothed}. Selective application of artificial viscosity techniques, such as the time dependent viscosity \citep{morris1997switch, dolag2005turbulent} and shock-indicator-based viscosity \citep{cullen2010inviscid} have been tried for counteracting this. 
\citet{sigalotti2008adaptive} improved the standard SPH near sharp discontinuities adopting adaptive density kernel estimation (ADKE) techniques. However, the parameters for ADKE have to be tuned to the problem at hand \citep{puri2014comparison}. Hence it is hard to implement ADKE in real applications.

Besides adding dissipation explicitly, \citet{inutsuka2002reformulation} proposed an innovative approach to introduce dissipation implicitly by reformulating the SPH convolution integrals. Artificial viscosity is introduced implicitly by using iterative solution to a Riemann problem at an imaginary interface between an interacting particles pair. This method, named as Godunov SPH (GSPH), is attractive because it eliminates parameterization and hence user intervention associated with artificial viscosity coefficients. The GSPH method has been shown to be able to handle shock discontinuities \citep{inutsuka2002reformulation, cha2003implementations,iwasaki2011smoothed, puri2014approximate}, introduces less damping to turbulent mixing \citep{cha2010kelvin, borgani2012hydrodynamic} than standard SPH, and ameliorate pressure ``blips" (or ``wiggle" called by others) at the contact discontinuity \citep{borgani2012hydrodynamic}. However, the amount of dissipation introduced by GSPH is stll excessive than needs. When using a linear interpolation of the volume function, it has been shown \citep{borgani2012hydrodynamic} that the excess of diffusion, which manifests itself in the shock tube tests as a smooth transition at the rarefaction fan, is such to prevent the development of the Kelvin-Helmholtz (KH) instability.
%The development of KH instability is also significantly affected by number of neighbouring particles for cubic interpolation of the volume function.

The random choice method (RCM) was introduced by \citet{glimm1965solutions} for the construction of solutions of systems of nonlinear hyperbolic conservation laws. \citet{chorin1976random} developed the RCM as a practical computational method for solutions to one dimensional (1D) Euler equations of gas dynamics. Chorin's work was followed by numerous researchers, with applications and further improvements \citep{sod1977numerical,concus1979numerical,colella1982glimm, freistuhler1992numerical}. In essence, the solution is advanced in time by a sequence of operations that include solving of Riemann problems and a stochastic sampling of the solution. See \citep{toro2013riemann} for details. When applied to one dimensional time dependent Euler equations \citep{colella1982glimm}, diffusive errors from spatial averaging are completely eliminated and hence discontinuities, such as shocks or material interface, can be resolved as true discontinuities.
For unsteady flows in genuine two (or higher) space dimensions, however, such very desirable property of RCM do not seem to persist \citep{colella1982glimm}.
%This need to be proven ---> But at least, if the good property of preserving true discontinuity is still valid in 1D shock tube problem, it should be preserved in multiple-dimensional implementations. Because in current framework of RSPH, no matter what is the dimension of the problem, the techniques is only implemented in a one-dimensional scenario.

In this investigation, we introduce a new method, which will be referred to as RSPH in later paragraph, by integrating RCM method within framework of SPH. 
The new scheme inherits the attraction of classical RCM, introducing less but sufficient artificial viscosity and causing less smearing of discontinuities than other methods for 1D problems. Implementing of RCM within the framework of SPH only involves sampling of solutions of Riemann problem in 1D local coordinate system no matter what is the dimensionality of the problem. That is to say, no multi-dimensional coordinate is involved even for multi-dimensional problems. Thereby, the desirable property of RCM method in 1D should be retained in higher dimensional implementations within the framework of SPH. This hypothesis is confirmed by a three-dimensional (3D) free jet flow simulation.

To summarize, RSPH has the following advantages over standard SPH:
\begin{itemize}
\item RSPH is able to capture discontinuities with less smearing.
\item The RSPH is able to adaptively adjust equivalent artificial viscosity coefficients, assigning larger artificial viscosity coefficients around the shock and small artificial viscosity coefficients otherwhere.
\item RSPH is able to alleviate pressure ``wiggle" at the contact discontinuity, a common issue in classical SPH.
\end{itemize}
Compared with GSPH, which is also based on solution of local Riemann problems, RSPH introduces less but sufficient dissipation than GSPH, especially in the region away from shocks. Such feature of RSPH is especially beneficial for simulating of scenarios involving turbulence mixing, for which excessive dissipation would suppress mixing. In addition, RSPH also shows faster convergence rate than GSPH.

The scheme of this paper is follows. We start with a brief review of standard SPH and GSPH in the second section. The RSPH is then explained in details in the third section followed by the numerical tests section, where both one 1D and 3D inviscid Euler equations (see Eq. (\ref{eq:gov-cs-rho}) - (\ref{eq:gov-cs-e}) are solved by standard SPH, GSPH and RSPH. Limitations and potential applications of RSPH are discussed in last section.

Our investigation implements RSPH scheme in solving inviscid compressile Euler equations as the application case.  
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot \left(\rho \textbf{v} \right) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot \left(\rho \textbf{v} \textbf{v} + p\textbf{I}\right) = 0 \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot \left[\left(\rho E + p \right)\textbf{v}\right] = 0 \label{eq:gov-cs-e}
\end{align}
where $\rho$ is the density, $\textbf{v}$ is the velocity, $\textbf{I}$ is a unit tensor.
$E = e + K $ is the total energy which is a summation of kinetic energy $K$ and internal energy $e$.
The system of equations is closed by equation of state (EOS) for idea gas.
\begin{equation}
p = \left(\gamma - 1\right)\rho e \label{eq:EOS}
\end{equation}
with $\gamma=1.4$.

For any SPH scheme, governing equations in Lagrange form are needed. By deducting kinetic energy from energy equation, subtracting mass conservation from momentum equation, combining transient term and advection term into material derivative (for any function $A$, material derivative is defined as $\frac{D A}{Dt} = \frac{\partial A}{\partial t} + \textbf{v} \cdot \nabla A$) term, the governing equations are converted into the Lagrange form. 
\begin{align}
\dfrac{D \rho}{D t} + \rho \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-rho}\\
\dfrac{D \textbf{v}}{D t} + \dfrac{\nabla P}{\rho} =0 \label{eq:gov-nc-v}\\
\dfrac{D e}{D t} + \dfrac{P \nabla \cdot \textbf{v}}{\rho} = 0 \label{eq:gov-nc-e}
\end{align}
%\cutet{puri2014comparison} propose an addition of an artificial heating term to the GSPH scheme to eliminate unphysical spikes in the thermal energy at the contact discontinuity.  --> It is worthwhile to check whether RSPH is able to avoid this phenomena without using his techniques.
%Another issue of RCM: the very desirable properties of RCM: no diffusive errors from spatial averaging do not seem to persist for flows in two space dimensions. For two dimensional steady supersonic flow, one of the space variables can be treated as a time-like one and RCM can be applied in a way similar to that for one-dimensional unsteady flow.
%--Then what about RSPH?

\section{Standard SPH and GSPH method}
The RSPH method is based on standard SPH and GSPH. In addition, these three methods will be compared in the numerical tests section. So it is necessary to review discretization of the compressible Euler equations by SPH and GSPH. We present here a brief review on standard SPH and GSPH providing corresponding discretized governing equations.

\subsection{Standard SPH method}
In SPH, the domain is discretized by a set of particles or discretization points and the position of each particle is updated at every time step based on the motion computed. Approximation of all field variables (velocity, density and pressure, ect.) is obtained by interpolation based on discretization points. The physical laws (such as conservation laws of mass, momentum and energy) written in the form of partial differential equations (PDEs) or ordinary differential equations (ODEs) need to be transformed into the Lagrangian particle formalism of SPH. Using a kernel function that provides the weighted estimation of the field variables at any point, the integral equations are estimated as sums over particles in a compact subdomain defined by the support of the kernel function associated with the discretization points. Thus, field variables associated to the particle are updated based on its neighbors. Each kernel function has a compact support determined by smoothing length of the particle. Various tricks can be used to conserve linear and angular momentum and thermal energy \citep{monaghan1992smoothed}. Special treatments are also needed for second order derivative terms \citep{monaghan2005smoothed}. We only refer here to one of these possible discretization of compressible Euler equations, while we refer to \citet{ monaghan2005smoothed, liu2010smoothed, price2012smoothed} for complete formal derivation and other discretization formulations.
\begin{align}
<\rho_a> = \sum_b m_b w_{ab} \left(h\right) \label{eq:ns-sph-d} \\
\left\langle\dfrac{d \textbf{v}_a}{d t}\right\rangle = -\sum_b m_b \left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \nabla_a w_{a b}\left(h\right) \label{eq:ns-sph-v} \\
\left\langle\dfrac{d e_a}{d t}\right\rangle=
 0.5\sum_b m_b \textbf{v}_{a b}\left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \cdot \nabla_a w_{a b}\left(h\right) \label{eq:ns-sph-e}
\end{align}
where, $a$ is the SPH particle index, $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$. The smoothing length $h$ is taken as $h=\frac{h_a + h_b}{2}$. $\Pi$ is an artificial viscosity term.
One of commonly used models of artificial viscosity \citep{monaghan1983shock} is:
\begin{equation}
\Pi_{ab}=- \frac{\nu}{\bar{\rho}_{ab}} \dfrac{ \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2}
\label{eq:art-vis-original}
\end{equation}
The coefficient $\nu$ is defined as:
\begin{equation}
\nu = \alpha \bar{h}_{ab} \bar{c}_{ab}
\end{equation}
where 
\begin{align}
\bar{c}_{ab} = \dfrac{c_a + c_b}{2} \\
\bar{\rho}_{ab} = \dfrac{\rho_a + \rho_b}{2} \\
\textbf{x}_{ab}=\textbf{x}_a-\textbf{x}_b
\end{align}
The artificial viscosity term $\Pi_{ab}$ is a Galilean invariant and vanishes for rigid rotation. It produces a repulsive force between two particles when they are approaching each other and an attractive force when they are receding from each other.
An extra term was added to $\nu$ considering aspects of the dissipative term in shock solutions based on Riemann solvers and lead to a new formulation of artificial viscosity \citep{monaghan1992smoothed}. We adopt this new formulation in our simulation:
\begin{equation}
\Pi_{ab}^{\beta} = 
\begin{cases} 
      \dfrac{- \alpha \mu_{ab} \bar{c}_{ab} + \beta \mu_{ab}^2} {\bar{\rho}_{ab}} & \textbf{v}_{ab} \cdot \textbf{x}_{ab} < 0\\
      0 & \textbf{v}_{ab} \cdot \textbf{x}_{ab} > 0
\end{cases}
\label{eq:art-vis-shock}
\end{equation}
where
\begin{equation}
\mu_{ab} = \dfrac{h \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2} 
\end{equation}
%$\alpha$ and $\beta$ are two parameters that are free to be adjusted for each case. 
$\alpha$ and $\beta$ are two parameters that can be adjusted for different cases.
$\alpha = 1$ and $\beta = 2$ are recommended \citep{monaghan2005smoothed} for best results and mostly adopted. In the numerical test section, these two parameters are varied to show the effect different amount of numerical dissipation on simulation results. $\eta$ is usually taken as 0.1 to prevent singularities.

>>>>

\modulolinenumbers[5]
\journal{Journal of Computational Physics}
%%%%%%%%%%%%%%%%%%%%%%%
%% Elsevier bibliography styles
%%%%%%%%%%%%%%%%%%%%%%%
%% To change the style, put a % in front of the second line of the current style and
%% remove the % from the second line of the style you would like to use.
%%%%%%%%%%%%%%%%%%%%%%%
%% Numbered
\bibliographystyle{model1-num-names}
%% Numbered without titles
%\bibliographystyle{model1a-num-names}
%% Harvard
%\bibliographystyle{model2-names.bst}\biboptions{authoryear}
%% Vancouver numbered
%\usepackage{numcompress}\bibliographystyle{model3-num-names}
%% Vancouver name/year
%\usepackage{numcompress}\bibliographystyle{model4-names}\biboptions{authoryear}
%% APA style
%\bibliographystyle{model5-names}\biboptions{authoryear}
%% AMA style
%\usepackage{numcompress}\bibliographystyle{model6-num-names}
%% `Elsevier LaTeX' style
%\bibliographystyle{elsarticle-num}
%%%%%%%%%%%%%%%%%%%%%%%
\begin{document}
\begin{frontmatter}
\title{IA Random Choice SPH Scheme with Adaptive Viscosity}
%\tnotetext[mytitlenote]{Can add  title note here}
%% Group authors per affiliation:
\author{Zhixuan Cao}
\author{Abani Patra \fnref{mycorrespondingauthor}}
\address{Department of Mechanical and Aerospace Engineering,The State University of New York at Buffalo, Amherst, NY, 14260-4200, USA}
\author{E. Bruce Pitman}
\address{Department of Materials Design and Innovation, The State University of New York at Buffalo, Amherst, NY, 14260-4200, USA}
%\fntext[myfootnote]{Since 1880.}
%% or include affiliations in footnotes:
%\author[mymainaddress,mysecondaryaddress]%{Elsevier Inc}
%\ead[url]{www.elsevier.com}
%\author[mysecondaryaddress]{Global Customer Service\corref{mycorrespondingauthor}}
\fntext[mycorrespondingauthor]{abani@buffalo.edu}
%\cortext[mycorrespondingauthor]{abani@buffalo.edu}
%\ead{abani@buffalo.edu}
%\address[mymainaddress]{1600 John F Kennedy Boulevard, Philadelphia}
%\address[mysecondaryaddress]{360 Park Avenue South, New York}

\begin{abstract}
The classical smoothed particle hydrodynamics (SPH) method
employs artificial viscosity to properly capture shock waves in hydrodynamics. This artificial viscosity typically
produces more dissipation than is needed to smooth shocks, incorrectly  smears 
contact discontinuities, and can overwhelm fluid turbulence. 
Several studies have proposed highly tuned versions of artificial viscosity,
turning on and off near shocks or other troublesome wave features. 
Unfortunately these schemes usually produce more dissipation than is appropriate.
A different kind of solver adapts Godunov\rq{}s idea of solving local Riemann problems 
as building blocks for a numerical SPH solver. Of course a first order Godunov solver still introduces an effective numerical diffusion that can infect the entire numerical solution.
We propose a novel SPH scheme that combines an approximate
 version of Glimm\rq{}s Random Choice method (RCM)
with SPH. RCM properly resolves and then samples all hydrodynamic waves, and
introduces no numerical diffusion into a simulation. Our version approximately resolves
hydrodynamic waves, and samples the approximate solution.
We test this method on standard one dimensional shock tube problems.
Our ultimate goal is to apply this method to jet flow. Because of the high shear rate
in this application, we introduce a highly tuned eddy viscosity that is applied only near 
the shear surface.  The combination of the SPH and RCM together with a shear viscosity
produces nicely resolved jet profiles.

\end{abstract}
\begin{keyword}
Smoothed Particle Hydrodynamics (SPH), Random Choice Method (RCM), Riemann Solver, Artificial viscosity, Hyperbolic PDEs
\end{keyword}
\end{frontmatter}

\linenumbers
\section{Introduction}
Smoothed particle hydrodynamics (SPH) \citep{gingold1977smoothed,lucy1977numerical} is a mesh-free particle method based on a Lagrangian formulation of the governing equations of  motion.
SPH has been widely applied to different areas in engineering and science. As
should be expected, SPH possesses certain advantages and certain disadvantages relative to finite difference and finite element solvers. Principally, as a Lagrangian solver SPH conserves mass 
exactly. 
In addition, because of its formulation SPH automatically concentrates numerical effort in regions of
high density and adapts to  arbitrary geometries. On the other hand, 
SPH has difficulty capturing shock waves and contact discontinuities. To overcome numerical
oscillations at discontinuities, classical SPH introduces
an artificial viscosity term. At it is most often used, when two SPH particles
are approaching, a von Neumann type viscosity is introduced
\citep{monaghan1983shock, monaghan1997sph,klapp2012strong}. 
This kind of viscosity, however, 
 usually provides more dissipation  than is truly needed at a discontinuity, and the excess
 dissipation ultimately
 corrupts the entire computation. Various methods have been introduced to tune 
 the artificial viscosity, turning it on and off for particular wave features. However, 
 tuning switches cannot properly accommodate every kind of wave that might arise, 
 and 
 spurious effects and unphysical damping persist (see, for example, \citet{borgani2012hydrodynamic,flebbe1994smoothed,morris1997switch, dolag2005turbulent,cullen2010inviscid,sigalotti2008adaptive,puri2014comparison}).

To overcome troublesome oscillations, \cite{inutsuka2002reformulation} and later \cite{cha2003implementations} formulated an SPH scheme that uses Riemann problems to 
resolve the interactions of particles. This Godunov-SPH (GSPH) method can resolve shocks and pressure 
oscillations without additional
artificial viscosity 
 \citep{inutsuka2002reformulation, cha2003implementations,iwasaki2011smoothed, puri2014approximate,borgani2012hydrodynamic}. It has also been shown to introduce less damping in turbulent mixing \citep{cha2010kelvin, borgani2012hydrodynamic} than standard SPH. 
However, the dissipation introduced by GSPH is still larger than physically relevant, to the point
of inhibiting the development of the Kelvin-Helmholtz instability under certain conditions
 \citep{borgani2012hydrodynamic}.

The Godunov approach to solving Riemann problems in the context of numerical methods for conservation laws is to 
integrate the Riemann problem solution over a spatial interval of interest. A different approach is given by
the 
random choice method (RCM) was introduced by \citet{glimm1965solutions} for the construction of solutions of systems of nonlinear hyperbolic conservation laws. The RCM probabilistically samples the
Riemann problem solution in each interval, and thus introduces no viscosity whatsoever. Because this
sampling interrogates the exact Riemann problem solution, no over- or under-shoots are introduced,
so, irrespective of the absence of viscosity, no non-physical oscillations are introduced either.

\citet{chorin1976random} developed the RCM into a practical computational method for solutions of the Euler equations in one space dimension.
This approach was later extended to two dimensions \cite{chorinporousmedium}. Others
further refined the random choice method \citep{sod1977numerical,concus1979numerical,colella1982glimm, freistuhler1992numerical,toro2013riemann}. 
%This need to be proven ---> But at least, if the good property of preserving true discontinuity is still valid in 1D shock tube problem, it should be preserved in multiple-dimensional implementations. Because in current framework of RSPH, no matter what is the dimension of the problem, the techniques is only implemented in a one-dimensional scenario.
%
Later, Harten and Lax \cite{hartenlax} proposed approximately solving a Riemann problem based 
on a finite difference scheme, and sampling this approximate solution. Our approach here 
is based on the Harten-Lax proposal. 
We introduce a new method which we dub RSPH, by randomly sampling the interaction
of pairs of SPH particles. 
This scheme produces dissipation-free solutions, as does classical RCM, but for a Lagrangian 
system.
 Because the 
interaction of SPH particles are inherently one-dimensional, the RCM framework
is straightforward to implement.
We test this RSPH idea first on classical shock tube problems, where we can compute very accurate
solutions.
We then implement RSPH for a three-dimensional free jet flow simulation, a prototype for
our ultimate goal of simulating a
Plinian volcanic eruption.

%To summarize, RSPH has the following advantages over standard SPH:
%\begin{itemize}
%\item RSPH is able to capture discontinuities with less smearing.
%\item The RSPH is able to adaptively adjust equivalent artificial viscosity coefficients, assigning larger artificial viscosity coefficients around the shock and small artificial viscosity coefficients otherwhere.
%\item RSPH is able to alleviate pressure ``wiggle" at the contact discontinuity, a common issue in classical SPH.
%\end{itemize}

Section \ref{sph} provides a brief review of standard SPH, and shows how the
GSPH method is defined in section \ref{sec:GSPH-method}. The RSPH method is defined in Section \ref{rsph}.
One dimensional tests are shown in Section \ref{tests}, followed by the jet flow
example in Sec. \ref{jet}. 
Other applications and limitations  of RSPH are discussed in Section \ref{discussion}

\section{SPH Formulations} \label{sph}
Throughout this paper we will be concerned with solving the Euler equations, written as
\begin{align}
\dfrac{\partial \rho}{\partial t} + \nabla \cdot \left(\rho \textbf{v} \right) = 0 \label{eq:gov-cs-rho} \\
\dfrac{\partial \rho \textbf{v}}{\partial t} + \nabla \cdot \left(\rho \textbf{v} \textbf{v} + p\textbf{I}\right) = 0 \label{eq:gov-cs-v} \\
\dfrac{\partial \rho E}{\partial t} + \nabla \cdot \left[\left(\rho E + p \right)\textbf{v}\right] = 0 \label{eq:gov-cs-e}
\end{align}
Here $\rho$ is the density, $\textbf{v}$ is the velocity, $\textbf{I}$ is a unit tensor.
$E = e + K $ is the total energy which is a sum of a 
kinetic energy $K$ and an internal energy $e$.
The system of equations is closed by the equation of state  for an idea gas
\begin{equation}
p = \left(\gamma - 1\right)\rho e \label{eq:EOS}
\end{equation}
with $\gamma=1.4$.

SPH is a Lagrangian scheme, so we rewrite the governing equations  as
\begin{align}   %\label{eq:lagrangian}
\dfrac{D \rho}{D t} + \rho \nabla \cdot \textbf{v} = 0 \label{eq:gov-nc-rho}\\
\dfrac{D \textbf{v}}{D t} + \dfrac{\nabla P}{\rho} =0 \label{eq:gov-nc-v}\\
\dfrac{D e}{D t} + \dfrac{P \nabla \cdot \textbf{v}}{\rho} = 0 \label{eq:gov-nc-e}
\end{align}
In this formulation we have written the material derivative  as 
$\frac{D A}{Dt} = \frac{\partial A}{\partial t} + \textbf{v} \cdot \nabla A$) for a conserved variable $A$.

%\cutet{puri2014comparison} propose an addition of an artificial heating term to the GSPH scheme to eliminate unphysical spikes in the thermal energy at the contact discontinuity.  --> It is worthwhile to check whether RSPH is able to avoid this phenomena without using his techniques.
%Another issue of RCM: the very desirable properties of RCM: no diffusive errors from spatial averaging do not seem to persist for flows in two space dimensions. For two dimensional steady supersonic flow, one of the space variables can be treated as a time-like one and RCM can be applied in a way similar to that for one-dimensional unsteady flow.
%--Then what about RSPH?

\subsection{Standard SPH}
In the classical SPH formulation a fluid is divided into   a number of particles whose 
location is updated at every time step. An approximation to the field variables (velocity, density and pressure) is obtained by interpolation of particle quantities using  a kernel.
We will only consider kernel functions with  compact support. Derivatives of variables are 
converted into derivatives of the kernel function, and different formulations of the
derivatives can be shown to possess various properties; see 
 \citep{monaghan1992smoothed,monaghan2005smoothed,liu2010smoothed, price2012smoothed} for details.
 Thus the density due to particle $a$ can be written as a sum
 $<\rho_a> = \sum_b m_b w_{ab} \left(h\right) \label{eq:ns-sph-d}$, where 
 where $m_b$ is the  mass of the $b^{th}$ SPH particle and $w_{ab}$ is 
 the kernel interpolating between 
 particles $a$ and $b$, with a kernel smoothing length $h$. 
 Particle position is updated at every time step according to  a simple advection
\begin{equation}
\left\langle\dfrac{d \textbf{x}_a}{dt}\right\rangle = \textbf{v}_a \label{eq:SPH-update-pos}
\end{equation}

 Using this framework, derivative terms in (\ref{eq:gov-nc-rho})--(\ref{eq:gov-nc-e}) are written 
\begin{align}
\left\langle\dfrac{d \textbf{v}_a}{d t}\right\rangle = -\sum_b m_b \left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \nabla_a w_{a b}\left(h\right) \label{eq:ns-sph-v} \\
\left\langle\dfrac{d e_a}{d t}\right\rangle=
 0.5\sum_b m_b \textbf{v}_{a b}\left(\dfrac{p_b}{\rho_b^2} + \dfrac{p_a}{\rho_a^2} + \Pi_{ab}\right) \cdot \nabla_a w_{a b}\left(h\right) \label{eq:ns-sph-e}
\end{align}
Here $\textbf{v}_{a b} = \textbf{v}_a - \textbf{v}_b$, the smoothing length $h$ 
is taken as $h=\frac{h_a + h_b}{2}$, and $\Pi$ is an artificial viscosity term.
One of commonly used models of artificial viscosity \citep{monaghan1983shock} is:
\begin{equation}
\Pi_{ab}=- \frac{\nu}{\bar{\rho}_{ab}} \dfrac{ \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2}
\label{eq:art-vis-original}
\end{equation}
where the coefficient $\nu$ is defined as
\begin{equation}
\nu = \alpha \bar{h}_{ab} \bar{c}_{ab}
\end{equation}
with
\begin{align}
\bar{c}_{ab} = \dfrac{c_a + c_b}{2} \\
\bar{\rho}_{ab} = \dfrac{\rho_a + \rho_b}{2} \\
\textbf{x}_{ab}=\textbf{x}_a-\textbf{x}_b
\end{align}
This form of the artificial viscosity  is a Galilean invariant and vanishes for rigid rotation. It produces a repulsive force between two particles when they are approaching each other and an attractive force when they are receding from each other.
Later,  Monaghan\citep{monaghan1992smoothed} added terms to the viscosity,
obtaining the form we will use in computations below:
\begin{equation}
\Pi_{ab}^{\beta} = 
\begin{cases} 
      \dfrac{- \alpha \mu_{ab} \bar{c}_{ab} + \beta \mu_{ab}^2} {\bar{\rho}_{ab}} & \textbf{v}_{ab} \cdot \textbf{x}_{ab} < 0\\
      0 & \textbf{v}_{ab} \cdot \textbf{x}_{ab} > 0
\end{cases}
\label{eq:art-vis-shock}
\end{equation}
where
\begin{equation}
\mu_{ab} = \dfrac{h \textbf{v}_{ab} \cdot \textbf{x}_{ab}}{\textbf{x}_{ab}^2 + \left(\eta h\right)^2} 
\end{equation}
%$\alpha$ and $\beta$ are two parameters that are free to be adjusted for each case. 
Here $\alpha$ and $\beta$ are free parameters that are adjusted as needed. 
\citep{monaghan2005smoothed}  recommends $\alpha = 1$ and $\beta = 2$ as
good fits for many computations.
 $\eta$ is usually taken as 0.1.

In Eq. (\ref{eq:ns-sph-d}) - Eq. (\ref{eq:ns-sph-e}), $w_{a b}\left(h\right)$ is the kernel
weighting function, expressed more completely as
 $w\left(\textbf{x}_a - \textbf{x}_b, h\right)$. 
Many  weighting functions have been proposed, but, for simplicity,  we 
employ a truncated Gaussian in $d$ dimensions
\begin{equation}
w\left(\textbf{x} - \textbf{x} \prime, h \right) = 
\begin{cases} 
      \dfrac{1}{\left(h \sqrt{\pi}\right)^d} exp \left[- \left(\dfrac{\textbf{x} - \textbf{x} \prime}{h} \right)^2 \right] &  \vert \textbf{x} - \textbf{x} \prime \vert \leq 3h\\
      0 & \text{Otherwise}
\end{cases}
\label{eq:SPH-kernel}
\end{equation}
Derivatives of $w$ above can be written as 
\begin{equation}
\nabla w\left(\textbf{x} - \textbf{x} \prime, h \right) = 
\begin{cases} 
      -2\left(\dfrac{\textbf{x} - \textbf{x} \prime}{h}\right) \dfrac{1}{\left(h \sqrt{\pi}\right)^d} exp \left[- \left(\dfrac{\textbf{x} - \textbf{x} \prime}{h}\right)^2 \right] &  \vert \textbf{x} - \textbf{x} \prime \vert \leq 3h\\
      0 & \text{Otherwise}
\end{cases}
\label{eq:SPH-kernel-gradient}
\end{equation}

To complete our specification of the classical SPH scheme we must decide on the time 
integration method. A forward Euler step with sufficiently small timestep suffices for purposes of the exposition
here.

\subsection{GSPH} \label{sec:GSPH-method}
Inutsuka \cite{inutsuka2002reformulation} recognized that the Godunov approach of 
solving local Riemann problems might obviate the need for artificial viscosity.  This sub-section 
describes this GSPH method. Our discussion  anticipates that of
Section \ref{sec:RSPH-method}, and sub-section \ref{sec:Picking-up-single-state}
 in particular, which contain details that are only briefly
introduced here.
